### 2023-09-05 doc/15yrsChange.Rmd
misc
Update 15yrsChange.Rmd
various updates

@@

@@

- temperate deciduous forests play a critical role in the global carbon cycle, accounting for a substantial portion of the global forest C sink
- The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century
- However, these models do not incorporate nuisance species -- i.e., species whose local abundance has been dramatically increased as a result of human activities and that are causing ecological harm. 
- Nuisance species are known to increase tree mortality (non-endemic pests and pathogens) and decrease recruitment (deer, non-endemic plants) throughout the (Eastern US) temperate deciduous biome, yet we don't know net effect on current and future C cycling
- Here, we use 15 years of detailed census data from a large forest dynamics plot with a deer exclosure to understand how nuisance species are affecting C cycling
- we show increased biomass mortality, reduced ∆AGB (outside deer exclosure), reduced AGB (outside deer exclosure)?
- if current trends continue, forest will continue to lose C sequestration capacity/ likely to be C source in future
- thus, C sequestration potential of temperate deciduous forests is overestimated

@@

@@

The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with deciduous forests representing a substantial portion of this, sequestering >300 Tg C yr^-1^ [@pugh_role_2019]. 
Although currently a C sink, the future of the biome remains uncertain.
The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century [@finzi_carbon_2020; @wu_uncertainty_2023; @refs], but with declining CO~2~ sequestration capacity. <!--I think this is true, but check/ find more refs.-->
However, ... [@wu_uncertainty_2023]
<!--Global C models generally agree that the biome's CO~2~ sequestration capacity is likely to decline [@ahlstrom_robustness_2012].-->
Yet, global C models predict an wide range of future trajectories of CO~2~ sequestration [@ahlstrom_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014], including the effects of nuisance species.

Nuisance species...

- often non-endemic
- white-tailed deer is important endemic nuisance species, over-abundant because of human influence

Tree mortality ...

@@

@@

Here, we use 15 years of detailed census data from a large forest dynamics plot including a 4-ha deer exclosure to test the following hypotheses:
(1) canopy tree mortality and associated biomass loss are increasing, in large part due to non-endemic nuisance species (pests & pathogens)
(2) growth and recruitment (outside deer exclosure) have not kept pace with tree mortality, resulting in declining AGB (or at least ∆AGB)
(3) C sequestration (∆AGB) has declined, and will decline further if current trends continue

@@

>>>>

Cameron Dow

Jen Jordan

@@

for submission to Anthromes joint special collection in New Phytologist/ Plants, People, Planet: https://www.newphytologist.org/news/view/332

@@

- temperate deciduous forests play a critical role in the global carbon cycle, accounting for a substantial portion of the global forest C sink
- The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century
- However, these models do not incorporate nuisance species -- i.e., species whose local abundance has been dramatically increased as a result of human activities and that are causing ecological harm. 
- Nuisance species are known to increase tree mortality (non-endemic pests and pathogens) and decrease recruitment (deer, non-endemic plants) throughout the (Eastern US) temperate deciduous biome, yet we don't know net effect on current and future C cycling
- Here, we use 15 years of detailed census data from a large forest dynamics plot with a deer exclosure to understand how nuisance species are affecting C cycling
- we show increased biomass mortality, reduced ∆AGB (only outside deer exclosure?), reduced AGB (only outside deer exclosure?)
- if current trends continue, forest will continue to lose C sequestration capacity/ likely to be C source in future
- thus, C sequestration potential of temperate deciduous forests is overestimated

@@

# Papers to read
- earth system models (pay special attention to predictions and uncertainty for eastern US)
  - @wu_uncertainty_2023. - do they say anything about nuisance species?
  - @ahlstrom_robustness_2012
- @miller_overabundant_2023
  - what were there metrics? would it be worth applying their metrics to our forest in order to classify by their system?

@@

The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with deciduous forests representing a substantial portion of this, sequestering >300 Tg C yr^-1^ [@pugh_role_2019]. 
Although currently a C sink, the future of the biome remains uncertain.
The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century [@finzi_carbon_2020; @wu_uncertainty_2023], albeit with declining CO~2~ sequestration capacity [@ahlström_robustness_2012]. 
Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlstrom_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014].
One mechanism that is not represented in global C models is the impact of nuisance species -- i.e., endemic or non-endemic species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@refs]. 
Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning/niche models that seek to project future forest distribution and carbon stocks [**?** @wu_uncertainty_2023], nor in carbon offset projects [**???** @refs].
This is problematic in that nuisance species -- including non-endemic insect pests and pathogens, non-endemic plants, and over-abundant herbivores -- are dramatically impacting carbon cycling in temperate deciduous forests around the world [@refs].

## (paragraph on tree mortality from non-endemic pests and pathogens)

@@

## (paragraph on recruitment failure because of deer and non-endemic plants)

- white-tailed deer (*Odocoileus virginianus* ) is important endemic nuisance species, over-abundant because of human influence

@mcgarvey_effects_2013

@@

Here, we use 15 years of detailed census data from a large forest dynamics plot including a 4-ha deer exclosure to test the following hypotheses:
(1) canopy tree mortality and associated biomass loss are increasing, in large part due to non-endemic nuisance species (pests & pathogens)
(2) growth and recruitment (outside deer exclosure) have not kept pace with tree mortality, resulting in declining AGB (-∆AGB)
(3) C stocks have declined (-∆AGB), and will decline further if current trends continue

@@

## Study site

## Data collection

## Analyses
